The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-03-14 paper/iphas-dr2.tex
Spell check and added citations

\documentclass[useAMS,usenatbib]{mn2e}
\usepackage{amssymb,amsmath}
\usepackage[pdftex]{graphicx}
\usepackage{longtable} 
%\usepackage{lscape}
\usepackage{float}
\usepackage{booktabs}
\usepackage{aas_macros}

@@

\author[G. Barentsen
et. al]{Geert Barentsen$^{1}$\thanks{E-mail:geert@barentsen.be},
H. J. Farnhill$^1$,
J. E. Drew$^1$,
E. A. Gonz$\acute{\rm{a}}$lez-Solares$^2$, \newauthor
R. Greimel$^3$,
M. J. Irwin$^2$,
B. Mizalski$^{99}$,
C. Ruhland$^1$,
P. Groot$^4$,
A. Mampaso$^5$, \newauthor
S. E. Sale$^6$,
M. J. Barlow,
R. L. M Corradi,
J. J. Drake,
J. Fabregat,
B. T. Gaensicke,
A. S. Hales,
J. Irwin,
C. Knigge,
D. J. Lennon,
J. R. Lewis,
M. Mohr-Smith,
R. A. H. Morris,
T. Naylor,
Q. A. Parker,
S. Phillipps,
R. Raddi,
P. Rodriguez-Gil,
L. Sabin,
D. Steeghs,
Y. C. Unruh,
K. Viironen,
J. S. Vink,
N. A. Walton,
N. J. Wright,
A. A. Zijlstra,
plus all those we forgot.
\newauthor\\
$^{1}$School of Physics, Astronomy \& Mathematics, University of Hertfordshire, College Lane, Hatfield, Hertfordshire, AL10 9AB, U.K.\\
$^{2}$Institute of Astronomy, Cambridge University, Madingley Road, Cambridge, CB3 OHA, U.K.\\
$^{3}$IGAM, Institute of Physics, University of Graz, Universit\"atsplatz 5, Graz, Austria.\\
$^{4}$Afdeling Sterrenkunde, Radboud Universiteit Nijmegen, Faculteit NWI, Postbus 9010, 6500 GL Nijmegen, The Netherlands.\\
$^{5}$Instituto de Astrof\'isica de Canarias, 38200 La Laguna, Tenerife, Spain.\\
$^{6}$Rudolf Peierls Centre for Theoretical Physics, Keble Road, Oxford, OX1 3NP, U.K.\\
$^{99}$Please add your current affiliation.\\
}

@@

\begin{abstract} % BACKGROUND, OBJECTIVE, METHODS, RESULTS, CONCLUSIONS
The INT/WFC Photometric H$\alpha$ Survey 
of the Northern Galactic Plane (IPHAS)
is a 1800 square degrees imaging survey
covering the entire northern Milky Way at $|b| < 5^\circ$
in the $r$, $i$ and \ha\ filters 
using the Wide Field Camera (WFC) 
on the 2.5-meter Isaac Newton Telescope (INT).
It is the first survey to offer comprehensive CCD photometry
of point sources across the Galactic Plane at visible wavelengths.
We present the first quality-controlled and
uniformly-calibrated source catalogue
to have been extracted from the images,
providing single-epoch photometry
for 219 million unique sources
across 92\% of the survey area.
The observations were carried out between 2003 and 2013
at a median seeing of 1.1 arcsec
to a depth of $r=21.2\pm 0.5$, $i=20.0\pm 0.3$ and \ha$=20.3\pm 0.3$
($5\sigma$ limits, Vega system).
We explain the data reduction 
and quality control procedures,
describe and test the new uniform calibration,
and detail the construction of the source catalogue
and its quality warning flags.
We find that the calibration
and the photometric repeatability are accurate to
$\sigma=0.03$~mag.
Finally, we demonstrate the ability of the 
catalogue's unique
$r$-$i$/$r$-\ha\ colour-colour diagram to
(i) characterise stellar populations and extinction regimes
towards different Galactic sightlines
and (ii) select candidate \ha\ emission-line objects.
The catalogue which accompanies this paper
provides the much-needed visible-light counterpart
to several infrared surveys of the Galactic Plane,
and provides images for some of the most
crowded regions to be faced by Gaia.
\end{abstract}

@@

Over the years it has become apparent that the legacy of IPHAS 
will extend beyond the traditional \ha\ applications 
of identifying emission-line stars and nebulae. 
Through the provision of $r$, $i$ broadband photometry 
alongside H$\alpha$ data,
IPHAS has created the opportunity 
to study Galactic Plane populations 
in a new way.
For example, the survey’s unique $r-H\alpha$ colour, 
when combined with $r-i$,
has been shown to provide simultaneous constraints 
on intrinsic stellar colour and interstellar extinction \citep{Drew2008}. 
This has opened the door 
to a wide range of Galactic science applications, 
including the mapping of extinction across the Plane in three dimensions
and the probabilistic inference of stellar properties
\citep{Sale2009, Sale2010, Giammanco2011, Sale2012, Barentsen2013}. 
In effect, the availability of narrowband H$\alpha$ 
alongside $r, i$ magnitudes
provides coarse spectral information for huge samples of stars 
which are otherwise too faint or numerous 
to be targeted by spectroscopic surveys.
For such science applications to succeed however, 
it is vital that the imaging data is transformed 
into a homogeneously calibrated photometric catalogue, 
in which quality problems 
and duplicate detections are flagged.

@@

In \S\ref{sec:observations} we start by recapitulating the key points
of the survey observing strategy.
In \S\ref{sec:reduction} we describe the data reduction
and quality control procedures.
In \S\ref{sec:calibration} we explain the uniform re-calibration
and test our results against the Sloan Digital Sky Survey (SDSS).
In \S\ref{sec:catalogue} we explain how the source catalogue was compiled.
In \S\ref{sec:discussion} we discuss the properties of the catalogue,
and finally in \S\ref{sec:demonstration} we demonstrate
the scientific exploitation of the colour/magnitude diagrams
using the new catalogue.
In \S\ref{sec:conclusions} we conclude and outline
our future ambitions.

@@

\begin{figure*}
        \includegraphics[width=1\linewidth]{./plots/footprint/footprint_small.png}
        \caption{Survey area showing the footprints
        of all the quality-approved IPHAS fields
        which have been included in this data release.
        The area covered by each field has been coloured black
        with a semi-transparent opacity of 20\%,
        such that regions where fields overlap are darker.
        The IPHAS strategy is to observe each field twice
        with a small offset,
        and hence the vast majority of the area 
        is covered twice (dominant gray colour).
        There are small overlaps between all the neighbouring fields
        which can be seen as a honeycomb-style pattern
        of dark gray lines across the survey area.
        Regions with incomplete data are apparent as white gaps (no data) 
        or in light gray (denoting that only the offset position is missing).}
        \label{fig:footprint}
\end{figure*}

Figure~\ref{fig:footprint} shows the footprint 
of the quality-approved observations included in this work. 
The fields which remain missing 
-- covering 8\% of the survey area --
are predominantly located towards the Galactic anti-center 
at $l > 120^o$.
Fields at these longitudes can only be accessed from La Palma 
in the months of November-December,
which is when the weather and seeing conditions are often poor
at the INT and observing attempts have failed repeatedly.
To enable the survey to be brought to completion, 
a decision was made recently to limit repeats in this area 
to individual fields requiring replacement,
i.e. fresh observations in all 3 filters may only be obtained 
at one of the two offset positions.  
The catalogue is structured such that it is clear 
where a contemporaneous observation of both halves of a field pair
is not available.

@@

Finally, an astrometric solution is determined
based on the 2MASS point source catalog \citep{Skrutskie2006},
which itself is calibrated 
in the International Celestial Reference System (ICRS).
A provisional photometric calibration is also provided 
based on the average zeropoint
determined from a set of standard stars observed in the same night.
Sources are classified morphologically
-- stellar, galaxy or noise --
based on the curve-of-growth determined
from measuring the source intensity in a series of growing apertures.
Finally, the resulting source detection tables are also stored 
in multi-extension FITS files.

@@

Surveys which observe identical stars at different epochs
can use the repeat measurements to ensure a uniform calibration.
For example, 2MASS attained its global calibration
by observing six standard fields each hour, 
allowing zeropoint variations to be tracked 
over very short timescales \citep{Nikolaev2000}.
Alternatively, the SDSS and PanSTARRS surveys could benefit
from revisiting regions in their footprint to 
carry out a so-called \emph{ubercalibration} procedure,
in which multiple measurements of stars at different epochs
are used to fit the calibration parameters
\citep[e.g.][]{Padmanabhan2008,Schlafly2012}.

@@

We have been able to benefit from the
AAVSO Photometric All-Sky Survey
(APASS; http://www.aavso.org/apass)
to bring the vast majority of the survey 
onto a uniform scale.
Since 2009,
APASS has been using two 20\,cm-astrographs
to survey the entire sky down to $\sim$17th magnitude
in five filters which include Sloan $r$ and $i$.
The most recent catalogue available 
at the time of preparing this work was APASS DR7,
which provides data across $\sim$half of the IPHAS footprint
The overlap regions are shown in Fig.~\ref{fig:apass_r}.
The photometric accuracy of APASS is currently estimated 
to be at the level of 3\% (Henden, private communication),
which is significantly better 
than the provisional calibration of IPHAS
for which we estimate the $1\sigma$-uncertainty to be $\sim$10\%.
APASS achieves its uniform accuracy 
by measuring each star at least two times in photometric conditions
-- along with ample standard fields --
using the large $3\times3$ square degrees field of view of its detectors.

@@

The resulting set of $\sim$220\,000 cross-matched stars were then used 
to derive APASS-to-IPHAS magnitude transformations
using a linear least-squares fitting routine, 
which iteratively removed $3\sigma$-outliers to improve the fit.
The solution converged to:
\begin{align} 
r_{\rm IPHAS} = r_{\rm APASS} - 0.121 + 0.032(r-i)_{\rm APASS} \label{eqn:apass_r} \\
i_{\rm IPHAS} = i_{\rm APASS} - 0.364 + 0.006(r-i)_{\rm APASS} \label{eqn:apass_i}
\end{align}
The root mean square (rms) residuals of these transformations 
are 0.041 and 0.051, respectively.
The small colour terms in the equations
indicate that the $r$ and $i$ filters 
are very similar in both surveys.
The transformations include a large fixed offset,
but this is simply due to the fact that 
APASS magnitudes are given in the AB system
and IPHAS uses magnitudes relative to Vega.
Separate transformations were derived for sightlines 
with varying extinction properties to investigate the robustness
of the transformations with respect to different reddening regimes.
The variations at these different
sightlines were found to be insignificant.
This is not surprising because heavily reddened objects 
are naturally scarce at $r<16$.

@@

\begin{table}
        \begin{center}
                \begin{tabular}{lcc}
                        \toprule
                         {\bf Before re-calibration} & Mean & $\sigma$  \\
                        \midrule
                        $r$ (IPHAS - APASS) & +0.014 & 0.104 \\
                        $i$ (IPHAS - APASS) & +0.007 & 0.108 \\
                        $r$ (IPHAS - SDSS) & +0.016 & 0.088 \\
                        $i$ (IPHAS - SDSS) & +0.010 & 0.089 \\
                        \bottomrule
                \end{tabular}
        \caption{Mean magnitude offsets for objects
                 crossmatched between IPHAS and APASS/SDSS
                 before the uniform re-calibration.
                 Eqns.~\ref{eqn:apass_r}-\ref{eqn:sdss_i} were applied
                 to the APASS/SDSS magnitudes to bring them into the
                 Vega-based IPHAS system prior to computing
                 the offsets.)
                 }
        \label{tbl:offsets_before}

@@

We then turned to the uniform calibration of the \ha\ data.
It is not possible to re-calibrate the narrowband 
in the same way as the broadbands,
because the APASS survey does not offer \ha\ photometry.
We can reasonably assume however,
that the corrections required for $r$ and \ha\ are identical
because the IPHAS data-taking pattern ensured 
that a field's \ha\ and $r$-band exposures
were taken at essentially the same time, 
separated only by the $\sim$30\,sec read-out time required by the WFC.
We have hence corrected the \ha\ zeropoints 
by re-using the zeropoint adjustements that were derived for the $r$ band
in the earlier steps.
An exception was made for 3\,101 fields
for which our quality-control routines revealed
zeropoint variations which exceeded 0.03~mag 
between consecutive fields,
suggesting strongly non-photometric conditions.
For good practice, the \ha\ zeropoints of these fields
were adjusted by solving Eqn.~\ref{eqn:matrix}
rather than linking them directly to the $r$-band shift.

@@

These global transformations were deemed adequate
for the purpose of validating our uniform calibration in a statistical sense.
Separate equations were derived towards different sightlines
to investigate the effects of varying reddening regimes.
The colour term was found 
to show some variation towards lowly reddened areas,
where red objects at $r-i > 1$ are rare.
The vast majority of red objects in the global sample
are those in highly reddened areas however,
which agree well with the global transformations
and dominate the statistical appraisal of our calibration.

@@

Information on the quality of each detection is included 
in a series of additional columns.
We draw attention to three useful flags
which warn about the likely presence of a systematic error:
\begin{enumerate}
\item The \emph{saturated} column is used to flag sources
for which the peak pixel height exceeds 55000 counts,
which is typically the case for stars brighter than 12-13th magnitude.
Although the pipeline attempts to extrapolate the brightness of
saturated stars based on the shape of the PSF,
such extrapolation is prone to error
and we do not recommend its use.
\item The \emph{deblend} column is used to flag sources 
which partially overlap with a nearby neighbour.
Although the pipeline applies a deblending procedure
to such objects, the procedure is currently applied separately
in each band and hence the $r$-$i$ and $r$-\ha\ colours
may be inaccurate if the deblending proceeded differently in each band.
\item The \emph{brightNeighb} column is used to flag sources which are located
within 5 arcmin from an object brighter than $V=7$ 
according to the Bright Star Catalogue (BSC; Hoffleit et al. 1991), 
or within 10 arcmin if the neighbour is brighter than $V=4$.
Such very bright stars are known to cause systematic errors
and spurious detections as a result of stray light 
and diffraction spikes.
\end{enumerate}
In addition to the above, we also created warning flags for internal bookkeeping.
For example, we flagged detections which fell in the strongly vignetted regions of the focal plane,
which were truncated by CCD edges,
or which were otherwise affected by bad pixels in the detector.
We will explain below that none of such detections 
have been included in the catalogue.
An alternative detection was available in essentially all these situations
as a result of the IPHAS field pair strategy,
and hence it was not necessary to include these internal warning flags
in the source catalogue.

@@

As the final step, the primary detections
selected above were compiled
into the 98-column source catalogue
that is described in Appendix~\ref{app:columns}
and is made available on our website and through Vizier.
The entire list of sources naturally includes 
a significant number of spurious entries
as a result of the very sensitive default detection settings
that are employed by the CASU pipeline.
To limit the size of the source catalogue,
we have decided to enforce three basic criteria
which must be met for a candidate source
to be included in the catalogue:
\begin{enumerate}
\item the source must have been detected at SNR$>5$ in at least
one of the bands, i.e. it is required that at least one of
\emph{rErr}, \emph{iErr} or \emph{haErr} is smaller
than 0.2 mag;
\item the shape of the source must not be an obvious
cosmic ray or noise artefact, i.e. we require
either \emph{pStar} or \emph{pGalaxy} to be
greater than 20\%;
\item the source must not have been detected in one of the strongly
vignetted corners of the detector, 
not have had any known bad pixels in the aperture,
and not have been on the edge of one of the CCDs,
i.e. we require the \emph{errBits} quality score
to be smaller than 64.
\end{enumerate}

@@

Firstly, Table~\ref{tab:reliable} details
a set of minimum quality criteria
which should benefit most applications
which require reliable $r-i$ and $r-$\ha\ colours
without removing more than 80\%
of the sources brighter than $r < 19$.
The listed criteria are designed to 
(i) remove low-S/N sources, 
(ii) remove saturated sources,
and (iii) remove objects for which the 2.3-arcsec diameter
aperture magnitude is inconsistent 
with its alternative 1.2-arcsec diameter measurement.
This last criterion is a proxy
for identifying objects which are affected
by inaccurate background subtraction
or failed source deblending.
A total of 86 out of 219 million sources 
(39\%) satisfy all the criteria listed in Table~\ref{tab:reliable}
and are hereafter referred to as ``reliable''.
For convenience, the catalogue contains a boolean column
named \emph{reliable} which flags these objects
and makes their selection easy.

@@

Figure~\ref{fig:uncertainties} shows the mean photometric
uncertainty as a function of magnitude in each band.
We find that the uncertainty typically
reaches 0.1~mag near 20.5 in $r$ 
and 19.5 in $i$/\ha\ for the default 2.3\arcsec\ aperture.
At this point we note that the average colour
across all sources in the catalogue is
$1.06\pm0.12$ for $r$-$i$ and $0.44\pm0.03$ for $r$-\ha.
The better depth of $r$ is hence compensated
by the fact that stars tend to have 
brighter magnitudes in both $i$ and \ha.

\begin{figure}
    \includegraphics[width=0.5\textwidth]{./plots/errors/uncertainties.pdf} 
    \caption{Mean photometric uncertainties
             for $r$ (top), $i$ (middle) and \ha\ (bottom).
             Data points shown are the average values of
             columns \emph{rErr}, \emph{iErr} and \emph{haErr}
             in the catalogue, 
             and the errorbars show the standard deviations.
             The dashed and solid lines indicate 
             the 10$\sigma$ and 5$\sigma$ limits, respectively.
             These uncertainties are based only on the (Poissonian)
             photon noise and hence this figure does not show
             systematic or calibration uncertainties.}
    \label{fig:uncertainties}
\end{figure}

>>>>

\documentclass[useAMS,usenatbib]{mn2e}
\usepackage{amssymb,amsmath}
\usepackage[pdftex]{graphicx}
\usepackage{longtable} 
\usepackage{float}
\usepackage{booktabs}
\usepackage{aas_macros}

@@

\author[G. Barentsen
et. al]{Geert Barentsen$^{1}$\thanks{E-mail:geert@barentsen.be},
H. J. Farnhill$^1$,
J. E. Drew$^1$,
E. A. Gonz$\acute{\rm{a}}$lez-Solares$^2$, \newauthor
R. Greimel$^3$,
M. J. Irwin$^2$,
B. Mizalski$^{99}$,
C. Ruhland$^1$,
P. Groot$^4$,
A. Mampaso$^5$, \newauthor
S. E. Sale$^6$,
M. J. Barlow$^{99}$,
R. L. M Corradi$^{99}$,
J. J. Drake$^{99}$,
J. Fabregat$^{99}$, \newauthor
B. T. Gaensicke$^{99}$,
A. S. Hales$^{99}$,
J. Irwin$^{99}$,
C. Knigge$^{99}$,
D. J. Lennon$^{99}$, \newauthor
J. R. Lewis$^{99}$,
M. Mohr-Smith$^{99}$,
R. A. H. Morris$^{99}$,
T. Naylor$^{99}$,
Q. A. Parker$^{99}$, \newauthor
S. Phillipps$^{99}$,
R. Raddi$^{99}$,
P. Rodriguez-Gil$^{99}$,
L. Sabin$^{99}$,
D. Steeghs$^{99}$, \newauthor
Y. C. Unruh$^{99}$,
K. Viironen$^{99}$,
J. S. Vink$^{99}$,
N. A. Walton$^{99}$,
N. J. Wright$^{99}$, \newauthor
A. A. Zijlstra$^{99}$,
plus those I forgot (sorry).
\newauthor\\
$^{1}$School of Physics, Astronomy \& Mathematics, University of Hertfordshire, College Lane, Hatfield, Hertfordshire, AL10 9AB, U.K.\\
$^{2}$Institute of Astronomy, Cambridge University, Madingley Road, Cambridge, CB3 OHA, U.K.\\
$^{3}$IGAM, Institute of Physics, University of Graz, Universit\"atsplatz 5, Graz, Austria.\\
$^{4}$Afdeling Sterrenkunde, Radboud Universiteit Nijmegen, Faculteit NWI, Postbus 9010, 6500 GL Nijmegen, The Netherlands.\\
$^{5}$Instituto de Astrof\'isica de Canarias, 38200 La Laguna, Tenerife, Spain.\\
$^{6}$Rudolf Peierls Centre for Theoretical Physics, Keble Road, Oxford, OX1 3NP, U.K.\\
$^{99}$Please add your current affiliation.\\
}

@@

\begin{abstract} % BACKGROUND, OBJECTIVE, METHODS, RESULTS, CONCLUSIONS
The INT/WFC Photometric H$\alpha$ Survey 
of the Northern Galactic Plane (IPHAS)
is a 1800 deg$^2$ imaging survey
covering the entire northern Milky Way at $|b| < 5^\circ$
in the $r$, $i$ and \ha\ filters 
using the Wide Field Camera (WFC) 
on the 2.5-meter Isaac Newton Telescope (INT) in La Palma.
We present the first quality-controlled and
uniformly-calibrated source catalogue
to have been extracted from the images,
providing single-epoch photometry
for 219 million unique sources
across 92\% of the survey area.
The observations were carried out between 2003 and 2013
at a median seeing of 1.1 arcsec
and to a depth of $r=21.2\pm 0.5$, $i=20.0\pm 0.3$ and \ha$=20.3\pm 0.3$
($5\sigma$ limits, Vega system).
We explain the data reduction 
and quality control procedures,
describe and test the new uniform calibration,
and detail the construction of the 97-column source catalogue.
We find that the calibration
and the photometric repeatability are accurate to
0.03~mag (rms),
and provide a series of recommended quality criteria
which can be used to select the most reliable data
from the catalogue.
Finally, we demonstrate the ability of the 
catalogue's unique
$(r-i)$/($r-\ha)$ colour-colour diagram to
(i) characterise stellar populations and extinction regimes
towards different Galactic sight-lines
and (ii) select candidate \ha\ emission-line objects.
IPHAS is the first survey to offer comprehensive CCD photometry
of point sources across the Galactic Plane at visible wavelengths,
providing the much-needed counterpart
to several existing near- and mid-infrared surveys,
and providing images for some of the most
crowded regions to be faced by Gaia.
\end{abstract}

@@

Over the years it has become apparent that the legacy of IPHAS 
will extend beyond the traditional \ha\ applications 
of identifying emission-line stars and nebulae. 
Through the provision of $r$, $i$ broadband photometry 
alongside H$\alpha$ data,
IPHAS has created the opportunity 
to study Galactic Plane populations 
in a new way.
For example, the survey’s unique $(r-\ha)$ colour, 
when combined with $(r-i)$,
has been shown to provide simultaneous constraints 
on intrinsic stellar colour and interstellar extinction \citep{Drew2008}. 
This has opened the door 
to a wide range of Galactic science applications, 
including the mapping of extinction across the Plane in three dimensions
and the probabilistic inference of stellar properties
\citep{Sale2009, Sale2010, Giammanco2011, Sale2012, Barentsen2013}. 
In effect, the availability of narrowband H$\alpha$ 
alongside $r, i$ magnitudes
provides coarse spectral information for huge samples of stars 
which are otherwise too faint or numerous 
to be targeted by spectroscopic surveys.
For such science applications to succeed however, 
it is vital that the imaging data is transformed 
into a homogeneously calibrated photometric catalogue, 
in which quality problems 
and duplicate detections are flagged.

@@

In \S\ref{sec:observations} we start by recapitulating the key points
of the survey observing strategy.
In \S\ref{sec:reduction} we describe the data reduction
and quality control procedures.
In \S\ref{sec:calibration} we explain the uniform re-calibration
and test our results against the Sloan Digital Sky Survey (SDSS).
In \S\ref{sec:catalogue} we explain how the source catalogue was compiled.
In \S\ref{sec:discussion} we discuss the properties of the catalogue
and in \S\ref{sec:demonstration} we demonstrate
the scientific exploitation of the colour/magnitude diagrams.
Finally, in \S\ref{sec:dataaccess} we discuss access
to the data and source code,
and in \S\ref{sec:conclusions} we conclude and outline
our future ambition

@@

\begin{figure*}
        \includegraphics[width=1\linewidth]{./plots/footprint/footprint_small.png}
        \caption{Survey area showing the footprints
        of all the quality-approved IPHAS fields
        which have been included in this data release.
        The area covered by each field has been coloured black
        with a semi-transparent opacity of 20\%,
        such that regions where fields overlap are darker.
        The IPHAS strategy is to observe each field twice
        with a small offset,
        and hence the vast majority of the area 
        is covered twice (dominant grey colour).
        There are small overlaps between all the neighbouring fields
        which can be seen as a honeycomb-style pattern
        of dark grey lines across the survey area.
        Regions with incomplete data are apparent as white gaps (no data) 
        or in light grey (denoting that only the offset position is missing).}
        \label{fig:footprint}
\end{figure*}

Figure~\ref{fig:footprint} shows the footprint 
of the quality-approved observations included in this work. 
The fields which remain missing 
-- covering 8\% of the survey area --
are predominantly located towards the Galactic anti-centre 
at $l > 120^o$.
Fields at these longitudes can only be accessed from La Palma 
in the months of November-December,
which is when the weather and seeing conditions are often poor
at the INT and observing attempts have failed repeatedly.
To enable the survey to be brought to completion, 
a decision was made recently to limit repeats in this area 
to individual fields requiring replacement,
i.e. fresh observations in all 3 filters may only be obtained 
at one of the two offset positions.  
The catalogue is structured such that it is clear 
where a contemporaneous observation of both halves of a field pair
is not available.

@@

Finally, an astrometric solution is determined
based on the 2MASS point source catalogue \citep{Skrutskie2006},
which itself is calibrated 
in the International Celestial Reference System (ICRS).
A provisional photometric calibration is also provided 
based on the average zeropoint
determined from a set of standard stars observed in the same night.
Sources are classified morphologically
-- stellar, galaxy or noise --
based on the curve-of-growth determined
from measuring the source intensity in a series of growing apertures.
Finally, the resulting source detection tables are also stored 
in multi-extension FITS files.

@@

Surveys which observe identical stars at different epochs
can use the repeat measurements to ensure a uniform calibration.
For example, 2MASS attained its global calibration
by observing six standard fields each hour, 
allowing zeropoint variations to be tracked 
over very short timescales \citep{Nikolaev2000}.
Alternatively, the SDSS and PanSTARRS surveys could benefit
from revisiting regions in their footprint to 
carry out a so-called \emph{ubercalibration} procedure,
in which multiple measurements of stars at different epochs
are used to fit the calibration parameters
\citep[e.g.][]{Ivezic2007,Padmanabhan2008,Schlafly2012}.

@@

We have been able to benefit from the
AAVSO Photometric All-Sky Survey
(APASS; http://www.aavso.org/apass)
to bring the vast majority of the survey 
onto a uniform scale.
Since 2009,
APASS has been using two 20\,cm-astrographs
to survey the entire sky down to $\sim$17th magnitude
in five filters which include Sloan $r$ and $i$ \citep{Henden2012}.
The most recent catalogue available 
at the time of preparing this work was APASS DR7,
which provides a good coverage across $\sim$half
of the IPHAS footprint.
The overlap regions are shown in Fig.~\ref{fig:apass_r}.
The photometric accuracy of APASS is currently estimated 
to be at the level of 3\% (Henden, private communication),
which is significantly better 
than the provisional calibration of IPHAS
for which we estimate the $1\sigma$-uncertainty to be $\sim$10\%.
APASS achieves its uniform accuracy 
by measuring each star at least two times in photometric conditions
-- along with ample standard fields --
using the large $3\times3$ square degrees field of view of its detectors.

@@

The resulting set of $\sim$220\,000 cross-matched stars were then used 
to derive APASS-to-IPHAS magnitude transformations
using a linear least-squares fitting routine, 
which iteratively removed $3\sigma$-outliers to improve the fit.
The solution converged to:
\begin{align} 
r_{\rm IPHAS} = r_{\rm APASS} - 0.121 + 0.032(r-i)_{\rm APASS} \label{eqn:apass_r} \\
i_{\rm IPHAS} = i_{\rm APASS} - 0.364 + 0.006(r-i)_{\rm APASS} \label{eqn:apass_i}
\end{align}
The root mean square (rms) residuals of these transformations 
are 0.041 and 0.051, respectively.
The small colour terms in the equations
indicate that the $r$ and $i$ filters 
are very similar in both surveys.
The transformations include a large fixed offset,
but this is simply due to the fact that 
APASS magnitudes are given in the AB system
and IPHAS uses magnitudes relative to Vega.
Separate transformations were derived for sight-lines 
with varying extinction properties to investigate the robustness
of the transformations with respect to different reddening regimes.
The variations at these different
sight-lines were found to be insignificant.
This is not surprising because heavily reddened objects 
are naturally scarce at $r<16$.

@@

\begin{table}
        \begin{center}
                \begin{tabular}{lcc}
                        \toprule
                         {\bf Before re-calibration} & Mean & $\sigma$  \\
                        \midrule
                        $r$ (IPHAS - APASS) & +0.014 & 0.104 \\
                        $i$ (IPHAS - APASS) & +0.007 & 0.108 \\
                        $r$ (IPHAS - SDSS) & +0.016 & 0.088 \\
                        $i$ (IPHAS - SDSS) & +0.010 & 0.089 \\
                        \bottomrule
                \end{tabular}
        \caption{Mean magnitude offsets for objects
                 cross-matched between IPHAS and APASS/SDSS
                 before the uniform re-calibration.
                 Eqns.~\ref{eqn:apass_r}-\ref{eqn:sdss_i} were applied
                 to the APASS/SDSS magnitudes to bring them into the
                 Vega-based IPHAS system prior to computing
                 the offsets.)
                 }
        \label{tbl:offsets_before}

@@

We then turned to the uniform calibration of the \ha\ data.
It is not possible to re-calibrate the narrowband 
in the same way as the broadbands,
because the APASS survey does not offer \ha\ photometry.
We can reasonably assume however,
that the corrections required for $r$ and \ha\ are identical
because the IPHAS data-taking pattern ensured 
that a field's \ha\ and $r$-band exposures
were taken at essentially the same time, 
separated only by the $\sim$30\,sec read-out time required by the WFC.
We have hence corrected the \ha\ zeropoints 
by re-using the zeropoint adjustments that were derived for the $r$ band
in the earlier steps.
An exception was made for 3\,101 fields
for which our quality-control routines revealed
zeropoint variations which exceeded 0.03~mag 
between consecutive fields,
suggesting strongly non-photometric conditions.
For good practice, the \ha\ zeropoints of these fields
were adjusted by solving Eqn.~\ref{eqn:matrix}
rather than linking them directly to the $r$-band shift.

@@

These global transformations were deemed adequate
for the purpose of validating our uniform calibration in a statistical sense.
Separate equations were derived towards different sight-lines
to investigate the effects of varying reddening regimes.
The colour term was found 
to show some variation towards lowly reddened areas,
where red objects at $r-i > 1$ are rare.
The vast majority of red objects in the global sample
are those in highly reddened areas however,
which agree well with the global transformations
and dominate the statistical appraisal of our calibration.

@@

Information on the quality of each detection is included 
in a series of additional columns.
We draw attention to three useful flags
which warn about the likely presence of a systematic error:
\begin{enumerate}
\item The \emph{saturated} column is used to flag sources
for which the peak pixel height exceeds 55000 counts,
which is typically the case for stars brighter than 12-13th magnitude.
Although the pipeline attempts to extrapolate the brightness of
saturated stars based on the shape of the PSF,
such extrapolation is prone to error
and we do not recommend its use.
\item The \emph{deblend} column is used to flag sources 
which partially overlap with a nearby neighbour.
Although the pipeline applies a deblending procedure
to such objects, the procedure is currently applied separately
in each band and hence the $(r-i)$ and $(r-\ha)$ colours
may be inaccurate if the deblending proceeded differently in each band.
\item The \emph{brightNeighb} column is used to flag sources which are located
within 5 arcmin from an object brighter than $V=7$ 
according to the Bright Star Catalogue (BSC; Hoffleit et al. 1991), 
or within 10 arcmin if the neighbour is brighter than $V=4$.
Such very bright stars are known to cause systematic errors
and spurious detections as a result of stray light 
and diffraction spikes.
\end{enumerate}
In addition to the above, we also created warning flags for internal bookkeeping.
For example, we flagged detections which fell in the strongly vignetted regions of the focal plane,
which were truncated by CCD edges,
or which were otherwise affected by bad pixels in the detector.
We will explain below that none of such detections 
have been included in the catalogue.
An alternative detection was available in essentially all these situations
as a result of the IPHAS field pair strategy,
and hence it was not necessary to include these internal warning flags
in the source catalogue.

@@

As the final step, the primary detections
selected above were compiled
into the 98-column source catalogue
that is described in Appendix~\ref{app:columns}
and is made available on our website and through Vizier.
The entire list of sources naturally includes 
a significant number of spurious entries
as a result of the very sensitive default detection settings
that are employed by the CASU pipeline.
To limit the size of the source catalogue,
we have decided to enforce three basic criteria
which must be met for a candidate source
to be included in the catalogue:
\begin{enumerate}
\item the source must have been detected at S/N$>5$ in at least
one of the bands, i.e. it is required that at least one of
\emph{rErr}, \emph{iErr} or \emph{haErr} is smaller
than 0.2 mag;
\item the shape of the source must not be an obvious
cosmic ray or noise artefact, i.e. we require
either \emph{pStar} or \emph{pGalaxy} to be
greater than 20\%;
\item the source must not have been detected in one of the strongly
vignetted corners of the detector, 
not have had any known bad pixels in the aperture,
and not have been on the edge of one of the CCDs,
i.e. we require the \emph{errBits} quality score
to be smaller than 64.
\end{enumerate}

@@

Firstly, Table~\ref{tab:reliable} details
a set of minimum quality criteria
which should benefit most applications
which require reliable $(r-i)$ and $(r-\ha)$ colours
without removing more than 80\%
of the sources brighter than $r < 19$.
The listed criteria are designed to 
(i) remove low-S/N sources, 
(ii) remove saturated sources,
and (iii) remove objects for which the 2.3-arcsec diameter
aperture magnitude is inconsistent 
with its alternative 1.2-arcsec diameter measurement.
This last criterion is a proxy
for identifying objects which are affected
by inaccurate background subtraction
or failed source deblending.
A total of 86 out of 219 million sources 
(39\%) satisfy all the criteria listed in Table~\ref{tab:reliable}
and are hereafter referred to as ``reliable''.
For convenience, the catalogue contains a boolean column
named \emph{reliable} which flags these objects
and makes their selection easy.

@@

Figure~\ref{fig:uncertainties} shows the mean photometric
uncertainty as a function of magnitude in each band.
We find that the uncertainty typically
reaches 0.1~mag near 20.5 in $r$ 
and 19.5 in $i$/\ha\ for the default 2.3\arcsec\ aperture.
At this point we note that the average colour
across all sources in the catalogue is
$1.06\pm0.12$ for $(r-i)$ and $0.44\pm0.03$ for $(r-\ha)$.
The better depth of $r$ is hence compensated
by the fact that stars tend to have 
brighter magnitudes in both $i$ and \ha.

\begin{figure}
    \includegraphics[width=0.5\textwidth]{./plots/errors/uncertainties.pdf} 
    \caption{Mean photometric uncertainties
             for $r$ (top), $i$ (middle) and \ha\ (bottom).
             Data points shown are the average values of
             columns \emph{rErr}, \emph{iErr} and \emph{haErr}
             in the catalogue, 
             and the error-bars show the standard deviations.
             The dashed and solid lines indicate 
             the 10$\sigma$ and 5$\sigma$ limits, respectively.
             These uncertainties are based only on the (Poissonian)
             photon noise and hence this figure does not show
             systematic or calibration uncertainties.}
    \label{fig:uncertainties}
\end{figure}
